---
title: "Assignment 2 - Universal Bank kNN Analysis"
author: "Hetu Patel"
date: '2026-10-01'
output: html_document
---
# Universal Bank k-NN Classification
# Business Problem - Universal Bank wants to increase the number of customers who accept personal loans. The purpose of this analysis is to use customer demographic and banking information to predict whether a customer will accept a personal loan offer.
# Solution - The classification method used in this assignment is k-Nearest Neighbors (k-NN). The Personal Loan variable is the target variable, where 1 indicates that the customer accepted the loan and 0 indicates that the customer did not accept the loan.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Installing required packages and calling them with library
# install.packages("dplyr")
# install.packages("class")
# install.packages("caret")
library(dplyr)
library(class)
library(caret)
```

```{r}
# loading the database and checking summary and data types present in it
bank_data <- read.csv("~/Desktop/Fall 2026/Fundamentals of machine Learning/Assignment 2/UniversalBank.csv")
summary(bank_data)
str(bank_data)
head(bank_data)
```
```{r}
# DATA PREPARATION
# Remove the ID and Zip Code columns from databse
updated_data <- bank_data[
  !names(bank_data) %in% c("ID", "ZIP.Code")
]

# converting education into factor and Creating dummy variables for education
updated_data$Education <- factor(updated_data$Education)
education_dummies <- dummyVars(
  ~ Education,
  data = updated_data,
  fullRank = FALSE
)

education_data <- predict(
  education_dummies,
  newdata = updated_data
)

education_data <- as.data.frame(education_data)

names(education_data) <- c(
  "Education_1",
  "Education_2",
  "Education_3"
)

# remove Education from updated_data and add the three dummy variales for education which was created
updated_data <- updated_data[
  !names(updated_data) %in% c("Education")
]
updated_data <- cbind(
  updated_data,
  education_data
)
names(updated_data)
head(updated_data)
```

```{r}
# Partitioning the data into training (60%) and validation (40%) sets.
set.seed(123)

train_index <- createDataPartition(
  updated_data$Personal.Loan,
  p = 0.60,
  list = FALSE
)

train_data <- updated_data[train_index, ]
validation_data <- updated_data[-train_index, ]
# Now the number of records for training and validation is as follows
# Training Set
nrow(train_data)
# Validation Set
nrow(validation_data)
```
```{r}
# Seperate predictors and target and scale the predictors
x_train <- train_data[
  !names(train_data) %in% "Personal.Loan"
]

y_train <- train_data$Personal.Loan

x_validation <- validation_data[
  !names(validation_data) %in% "Personal.Loan"
]

y_validation <- validation_data$Personal.Loan

preprocess <- preProcess(
  x_train,
  method = c("center", "scale")
)

x_train_scaled <- predict(
  preprocess,
  x_train
)

x_validation_scaled <- predict(
  preprocess,
  x_validation
)
```

```{r}
# Question 1 Classification using k=1. The customer is classified as [1 / 0]. 
# 1 = customer is predicted to accept the personal loan. 
# 0 = customer is predicted not to accept the personal loan.
new_customer <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1,
  Education_1 = 0,
  Education_2 = 1,
  Education_3 = 0
)

# Put variables in same order as training data
new_customer <- new_customer[
  , names(x_train)
]

# Scale new customer
new_customer_scaled <- predict(
  preprocess,
  new_customer
)

# k = 1
customer_k1 <- knn(
  train = x_train_scaled,
  test = new_customer_scaled,
  cl = y_train,
  k = 1
)

customer_k1
# Here customer was classified as 0 that is the customer is predicted not to accept the loan
```

```{r}
# QUESTION 2 - Finding Best k

k_values <- c(
  1, 3, 5, 7, 9,
  11, 13, 15, 17,
  19, 21
)

accuracy <- numeric(
  length(k_values)
)

for (i in seq_along(k_values)) {
  
  prediction <- knn(
    train = x_train_scaled,
    test = x_validation_scaled,
    cl = y_train,
    k = k_values[i]
  )
  
  accuracy[i] <- mean(
    prediction == y_validation
  )
}

k_results <- data.frame(
  k = k_values,
  Accuracy = accuracy
)

k_results

# Select best k
best_k <- k_results$k[
  which.max(k_results$Accuracy)
]
# The best k is [best_k], as it provides the best validation performance while balancing overfitting and underfitting. The results says the besk k is 1.
best_k

# Plot
plot(
  k_results$k,
  k_results$Accuracy,
  type = "b",
  xlab = "k",
  ylab = "Validation Accuracy",
  main = "Validation Accuracy for Different Values of k"
)

```

```{r}
# QUESTION 3 - The confusion matrix for the validation data using the best k.
# Make target variables factors with the same levels
y_train <- factor(as.character(y_train), levels = c("0", "1"))
y_validation <- factor(as.character(y_validation), levels = c("0", "1"))

# k-NN prediction
best_prediction <- knn(
  train = x_train_scaled,
  test = x_validation_scaled,
  cl = y_train,
  k = best_k
)

# Make prediction a factor with the same levels
best_prediction <- factor(
  as.character(best_prediction),
  levels = c("0", "1")
)

# Confusion matrix
confusionMatrix(
  best_prediction,
  y_validation,
  positive = "1"
)
```

```{r}
# QUESTION 4 - # Q4: Using the best k, the customer is classified as [1/0] for personal loan acceptance.
customer_best_k <- knn(
  train = x_train_scaled,
  test = new_customer_scaled,
  cl = y_train,
  k = best_k
)

customer_best_k

```

```{r}
# QUESTION 5 - Repartitioning  the data, this time into training, validation, and test sets (50% : 30% : 20%).
# Factor the personal loan
updated_data$Personal.Loan <- factor(
  updated_data$Personal.Loan,
  levels = c(0, 1)
)
set.seed(123)

# 50% Training
train_index2 <- createDataPartition(
  updated_data$Personal.Loan,
  p = 0.50,
  list = FALSE
)
train_data2 <- updated_data[train_index2, ]
remaining_data <- updated_data[-train_index2, ]

# 30% Validation and 20% Test
validation_index2 <- createDataPartition(
  remaining_data$Personal.Loan,
  p = 0.60,
  list = FALSE
)
validation_data2 <- remaining_data[
  validation_index2,
]
test_data2 <- remaining_data[
  -validation_index2,
]

# Check sizes
nrow(train_data2)
nrow(validation_data2)
nrow(test_data2)

# Separate predictors and target
x_train2 <- train_data2[
  !names(train_data2) %in% "Personal.Loan"
]
y_train2 <- train_data2$Personal.Loan
x_validation2 <- validation_data2[
  !names(validation_data2) %in% "Personal.Loan"
]
y_validation2 <- validation_data2$Personal.Loan
x_test2 <- test_data2[
  !names(test_data2) %in% "Personal.Loan"
]
y_test2 <- test_data2$Personal.Loan

# Scale using training data
preprocess2 <- preProcess(
  x_train2,
  method = c("center", "scale")
)

x_train2_scaled <- predict(
  preprocess2,
  x_train2
)

x_validation2_scaled <- predict(
  preprocess2,
  x_validation2
)

x_test2_scaled <- predict(
  preprocess2,
  x_test2
)

# Training Confusion Matrix
train_prediction2 <- knn(
  train = x_train2_scaled,
  test = x_train2_scaled,
  cl = y_train2,
  k = best_k
)

confusionMatrix(
  train_prediction2,
  y_train2,
  positive = "1"
)

# Validation Confusion Matrix
validation_prediction2 <- knn(
  train = x_train2_scaled,
  test = x_validation2_scaled,
  cl = y_train2,
  k = best_k
)

confusionMatrix(
  validation_prediction2,
  y_validation2,
  positive = "1"
)

# Test Confusion Matrix
test_prediction2 <- knn(
  train = x_train2_scaled,
  test = x_test2_scaled,
  cl = y_train2,
  k = best_k
)

confusionMatrix(
  test_prediction2,
  y_test2,
  positive = "1"
)
# The training, validation, and test results may differ because the training data is used to build the model, while validation and test data contain unseen observations.
```
 